### 2020-10-07 paper.Rmd
working on problem formalization

---
title             : "A bicriterion extension of k-means anticlustering to optimize between-group similarity"
shorttitle        : "Bicriterion K-means anticlustering"

@@

For the purpose of problem formalization, I adopt the notation Steinley 
(2006) provided in his comprehensive synthesis of the literature 
on k-means. K-means clustering is used to partition two-way, two-mode 
data (that is, $N$ elements each having $P$ attributes) into $K$ groups 
$(C_1., \ldots, C_K)$ where $C_k$ is the set of $n_k$ objects in group 
$k$. The partitioning is subject to the requirements that each element 
is assigned to exactly one group, formalized by the two constraints

@@

In this paper, I will discuss only such applications where groups are of 
equal size.

For a data matrix $\mathbf{X}_{N \times P} = \{x_{ij}\}_{N \times P}$, k-means 
clustering aims to minimize the error sum of squares (SSE), also known as the 
within-group variance. The SSE is the sum of the squared Euclidean distances 
between each data point and its cluster centroid. The cluster centroid is given 
by averaging the values on each variable over the elements within that cluster, 
where the centroid $\overline{x}_j ^{(k)}$ of the *j*th variable in the *k*th cluster is

@@

The SSE is then given as:

@@

Note there is a direct connection between $SSE$ and the cluster 
centroids [@spath1986]. If the complete centroid vector for cluster 
$C_k$ is given by

$$
\overline{\mathbf{x}}^{(k)} = 
  (\overline{x}_1^{(k)}, \overline{x}_2^{(k)} \ldots, \overline{x}_P^{(k)})
$$

and the $SSE$ is minimal, the centroid values 
$\overline{\mathbf{x}}^{(k)} (k = 1, \ldots, K)$ are as far away as 
possible from the overall data centroid $\overline{\mathbf{x}} = 
(\overline{\mathbf{x}}_1, \overline{\mathbf{x}}_2, \ldots, 
\overline{\mathbf{x}}_P$), where $\overline{\mathbf{x}}_p$ is the 
overall mean value on the $p$th attribute:

@@

If the $SSE$ is maximal, the cluster centroids are as close as possible to the 
overall centroid $\overline{\mathbf{x}}$ and therefore to each other 
[@spath1986]. Thus, k-means anticlustering directly optimizes the 
similarity of the mean values of the attributes across the different clusters.

## K-Variances

K-means anticlustering, i.e., maximizing SSE minimizes differences between the 
means of the input variables. I present a rather simple extension of the SSE 
criterion to simultaneously minimize differences with regard to the variances 
of the input variables. The variance is defined as the sum of the squared 
distances between all

The squared distance between an element $x = (x_1, \ldots, x_M)$ and the 
average of all elements $\overline{\mathrm{\boldsymbol{x}}}$ is:

$$
\vert \vert x - \overline{\mathrm{\boldsymbol{x}}} \vert \vert^2 = 
  \sum\limits_{m = 1}^{M}(x_{m} - \overline{\mathrm{\boldsymbol{x_m}}})^2
$$

Now define new data points representing the squared distances between 
data points and the grand mean:

$$ 
y_i = (x_i - \mathrm{\boldsymbol{x}})^2, \forall i \in N
$$

where $\mathrm{\boldsymbol{x}}$ is a vector of length $M$ containing 
the average for each feature.

Now:

$$
\sum\limits_{m = 1}^{M}(x_{m} - \overline{\mathrm{\boldsymbol{x_m}}})^2 + 
  \sum\limits_{m = 1}^{M}(y_{m} - \overline{\mathrm{\boldsymbol{y_m}}})^2 =
  \sum\limits_{m = 1}^{M}(x_{m} - \overline{\mathrm{\boldsymbol{x_m}}})^2 + 
    (y_{m} - \overline{\mathrm{\boldsymbol{y_m}}})^2
$$

If we define $x_{m + 1} = y_1$, $x_{m + 2} = y_2$, ..., $x_{2M} = y_M$.
this is the same as:

$$
\sum\limits_{m = 1}^{2M}(x_{m} - \overline{\mathrm{\boldsymbol{x_m}}})^2 
$$

Thus, we redefine the data points and include an additional $M$ features 
for each element. Then, we maximize $\mathit{Var}_{\mathrm{within}}$, 
which corresponds to a tradeoff between minimizing differences in mean 
features between clusters, and the average squared distance to the grand 
mean, i.e., the within-cluster variance. That is: We try to maximize the 
within-cluster variance, but to a certain degree we also strive to 
obtain similar within-cluster variance in each cluster.

@@

>>>>

---
title             : "A bicriterion extension of k-means anticlustering"
shorttitle        : "Bicriterion K-means anticlustering"

@@

For the purpose of problem formalization, I adopt the notation Steinley 
(2006) provided in his comprehensive synthesis of half a century of research on 
k-means. K-means clustering is used to partition two-way, two-mode 
data. That is, $N$ elements each having $P$ attributes into $K$ groups 
$(C_1., \ldots, C_K)$, $C_k$ being the set of $n_k$ objects in group 
$k$. The partitioning is subject to the requirements that each element 
is assigned to exactly one group, formalized by the two constraints

@@

For a data matrix $\mathbf{X}_{N \times P} = \{x_{ij}\}_{N \times P}$, k-means 
clustering aims to minimize the error sum of squares (SSE), also known as the 
within-group variance [@spath1986]. The SSE is the sum of the squared Euclidean 
distances between each data point and its cluster centroid; the *k*'th centroid 
is a vector of length $P$, where each entry is the mean value of one of the $P$ 
attributes across the observations belonging the *k*'th cluster ($k = 1, \ldots, 
K$). The centroid of the *k*'th cluster has the notation 
$\overline{\mathbf{x}}^{(k)}$ with
$$
\overline{\mathbf{x}}^{(k)} = 
  (\overline{x}_1^{(k)}, \overline{x}_2^{(k)} \ldots, \overline{x}_P^{(k)})
$$

The mean $\overline{x}_j ^{(k)}$ of the *j*th variable on the *k*th cluster is 
computed as:

@@

Based on the original data and the cluster centroids, the SSE---i.e., the k-means 
objective function---is given as:

@@

Note there is a direct connection between the $SSE$ and the location of the 
cluster centroids [@spath1986]. If the $SSE$ is minimal, the centroid values 
$\overline{\mathbf{x}}^{(k)} (k = 1, \ldots, K)$ are as far away as 
possible from the overall data centroid $\overline{\mathbf{x}} = 
(\overline{\mathbf{x}}_1, \overline{\mathbf{x}}_2, \ldots, 
\overline{\mathbf{x}}_P$), where $\overline{\mathbf{x}}_p$ is the 
overall mean value on the $p$th attribute:

@@

If the $SSE$ is maximal, the cluster centroids are as close as possible to the 
overall centroid $\overline{\mathbf{x}}$ and therefore to each other 
[@spath1986]. Thus, k-means anticlustering directly optimizes the 
similarity of the mean values of the $P$ attributes across the $K$ clusters.

## Bi-criterion extension

K-means anticlustering---i.e., maximizing the SSE---minimizes differences 
between the means of the input variables. In the following, a rather 
straight-forward extension of the SSE criterion is formulated that can be used 
to minimize differences with regard to the variances of the input variables at 
the same time.

For the sake of simplicity, we first consider a set of $N$ one-dimensional 
observations $(x_1, \ldots, x_N)$. The extension to the general case of $P$ 
attributes will be straight forward, because the formulation of the new 
criterion effectively reduces to the generation of new variables, leaving the 
k-means criterion SSE per sé unchanged.

The sample variance of this set of 
observations is computed as the mean of the squared distances between 
observations and their mean $\mathbf{x}$:

$$
\mathit{Var(x)} = \frac{1}{N} \sum\limits_{i=1}^{N} (x_i - \overline{\mathbf{x}})^2
$$

We next define a new variable $y = (y_1, \ldots, y_N)$, representing the squared 
deviation of each observation from the mean:

$$
y_i = (x_i - \overline{\mathbf{x}})^2
$$

The sample variance of $x$ is consequently the mean of these new values:

$$
\mathit{Var(x)} = \frac{1}{N} \sum\limits_{i=1}^{N} y_i
$$

Note at this point that the sample variance is basically defined as a mean 
value. We can use this insight and employ k-means anticlustering to equalize 
this *mean* across groups. For a given $K$, groups can be created that are 
similar with regard to the means of $y$---i.e., the variance of $x$---by 
maximizing the SSE, in this one-dimensional case given as:

$$
\mathit{SSE} = 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{i} - \overline{y} ^{(k)})^2
$$

## Weighted-sum approach

$$
\mathit{SSE_{comb}} = 
  w_1
  \sum\limits_{k=1}^{K} 
    \sum\limits_{i \in C_k}
    (x_{i} - \overline{x} ^{(k)})^2 + 
  w_2
  \sum\limits_{k=1}^{K} 
    \sum\limits_{i \in C_k}
    (y_{i} - \overline{y} ^{(k)})^2
$$

The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of 
equalizing means and variances, respectively.

## Unweighted case

Some simplifications can be arranged when ignoring the weights on the two 
criteria.

$$
\mathit{SSE_{comb}} = 
  \sum\limits_{k=1}^{K} 
    \sum\limits_{i \in C_k}
    (x_{i} - \overline{x} ^{(k)})^2 + 
  \sum\limits_{k=1}^{K} 
    \sum\limits_{i \in C_k}
    (y_{i} - \overline{y} ^{(k)})^2 \\
  = \sum\limits_{k=1}^{K} 
     \sum\limits_{i \in C_k}
     ((x_{i} - \overline{x} ^{(k)})^2 + 
     (y_{i} - \overline{y} ^{(k)})^2)
$$

When defining $z = (x, y)$, the SSE reduces to the standard case in (X):

$$
\mathit{SSE} = 
  \sum\limits_{j=1}^{2} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (z_{ij} - \overline{z}_j ^{(k)})^2
$$

This logic can be extended to the general case of $P$ attributes; for each 
attribute, a new variable is created as described above, and the entirety of 
all $2P$ variables is then used as input to standard k-means anticlustering.

@@

- extending on covariances?
